Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-06-16 output/paper/paper.tex
Updates from Overleaf

\begin{document}
    \maketitle
    \begin{abstract}
        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\section{Pandemic response demands timely data}
The new corona virus pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}. Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers \citep{Jajosky2004_evaluation_reporting}. The latest data are often not finalized, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

@@

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency\'s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths. Not how the model predicts additional deaths above the moving average line.

@@

\section{Implications and limitations}
The model proposed here can estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden. To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks. In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data. The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic. It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.

@@

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner. Given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure than Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so.

Moreover, deaths are reported from two distinct populations that seem to follow different trends. At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths. But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes. However, we have noted a clear decline in proportions of deaths reported the two first working days. For example the number of deaths occurring at the second of April $\approx 30\%$ of deaths where reported within the first two working days whereas for the eighteens of May only $\approx 10\%$ where reported during the two first working days. We address this by assuming that the deaths reported during the two first working days comes from a different population then the remainder of days.

>>>>

\begin{document}
    \maketitle
    \begin{abstract}
        The new coronavirus disease --- COVID-2019 --- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays. The methodology takes into account the observed distribution pattern of the lag. It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
	\end{abstract}
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\section{Pandemic response demands timely data}
The SARS-CoV-2 pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}. Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers \citep{Jajosky2004_evaluation_reporting}. The latest data are often not finalized, but change as new information is reported. In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.

@@

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency\'s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths. Note how the model predicts additional deaths above the moving average line.

@@

\section{Implications and limitations}
The model proposed here can estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden. To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks. In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data. The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the SARS-CoV-2 virus pandemic. It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.

@@

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner. Given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a differewnt reporting structure than Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so.

Moreover, deaths seem to be reported from two distinct populations that follow different trends. At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths. But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes. However, we have noted a clear decline in proportions of deaths reported the two first working days. For example, out of all deaths that occurred on the second of April, $\approx 30\%$ were reported within the first two workdays, whereas for the eighteenth of May only $\approx 10\%$ were reported that fast. We address this by assuming that the deaths reported during the two first working days comes from a different population than the remainder of days.
